Homicide Increases in the COVID-19 Era

### 2025-04-16 Draft4_2.qmd
final slides version and PDF

We will also investigate the relationship between homicide and enforcement factors. In particular, we will look at the LAPD budget in comparison to the homicide rate to determine if increased police funding is effective in reducing the number of homicides.

```{r DATA, output=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses
library(dplyr)
library(RSQLite)

@@

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.align="center", fig.cap="Crime densities in Los Angeles by LAPD Division, 2010-2019 and 2020-Present"}
query <- dbSendQuery(con, "Select * From crime_la Where year < 2020")
crime_la_2010_2019 <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}
query <- dbSendQuery(con, "Select Count(*) as Crimes, YearWeek From crime_la Group By YearWeek Order By YearWeek")
weekly_counts <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r, echo=FALSE, fig.cap = "Los Angeles Crimes by Day of the Week", warning=FALSE, message=FALSE, fig.align = "center"}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la <- rbind(crime_la_2010_2019, crime_la_2020_present)
# Extract the day of the week
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Percent of crime victims by gender, 2010 to present", fig.align = "center"}
query <- dbSendQuery(con, "SELECT COUNT(*) As count, Modified_Vict_Sex As `Vict.Sex` FROM (
    SELECT *,  
        CASE 
            WHEN `Vict.Sex` NOT IN ('M', 'F') THEN 'N'
            ELSE `Vict.Sex`
        END AS `Modified_Vict_Sex`
    FROM crime_la c
    WHERE `Vict.Sex` != '' AND `Vict.Sex` IS NOT NULL) s
    GROUP BY Modified_Vict_Sex
")
gender_counts <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}
zcta <- read_sf(dsn = "0_Data/Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
query <- dbSendQuery(con, "SELECT * FROM zctas_in_la")
LA_zctas <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Median household income by year and by ZIP Code Tabulation Area (ZCTA) in Los Angeles", fig.align = "center"}
# quantiles for the entire dataset (from 2018 to 2023) to divide the median income into groups
qntls <- round(quantile(zcta_no_NAs$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap="Annual Median Income and Homicide Count by ZCTA in Los Angeles with top 1% income and top 1% homicide rates labelled", fig.align = "center"}
zcta <- read_sf(dsn = "0_Data/Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
query <- dbSendQuery(con, "SELECT i.*, c.*
FROM cali_median_income i
INNER JOIN crimes_with_zips c
ON i.year = c.year AND CAST(i.ZCTA AS int) = c.ZCTA5CE20")
crime_income <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Monthly Frequency of Evictions and Crimes in the City of Los Angeles, February 2023 to February 2025"}

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Annual Number of Crimes by Los Angeles Police Department Bureau"}
crime_la$year <- lubridate::year(crime_la$DateParsed)
crime_by_area <- crime_la %>% group_by(AREA.NAME, year) %>% summarize(numberOfCrimes = n()) %>% ungroup()

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Strength of Correlations Between the Number of Crimes and Different Law Enforcement and Socioeconomic Factors"}
library(lubridate)
# creating dataset for correlations

@@

correlations_df <- crime_la_monthly %>% inner_join(unemployment_lapd_budget_evictions, join_by("year" == "Year", "month" == "month")) %>% dplyr::select(month, year, num_evictions, unemployment_rate, num_crimes, police_budget)

@@

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Homicide Victims by Age and Gender in Los Angeles 2010-Present"}
homicide_by_gender <- crime_la %>% filter(Crm.Cd.Desc == "CRIMINAL HOMICIDE") %>% dplyr::select(Vict.Sex, DateParsed, Vict.Age) %>% mutate(gender = factor(Vict.Sex, levels  = c("F", "M", "X"), labels = c("Female", "Male", "Other"))) %>% drop_na()

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Frequency of Crimes in Los Angeles Throughout the Day"}
crime_la$time <- sprintf("%04d",crime_la$TIME.OCC)
crime_la$Hour <- as.numeric(substr(as.character(crime_la$time), start = 1, stop = 2))

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Time series analyzing homicide rate trends over time by median income quartiles"}
query <- dbSendQuery(con, "
  SELECT i.S1903_C03_001E, x.year, i.ZCTA, x.LAT, x.LON
  FROM (
    SELECT c.LAT, c.LON, CAST(z.ZCTA5CE20 AS INTEGER) AS ZCTA5CE20, CAST(c.year AS INTEGER) AS year
    FROM crime_la AS c
    INNER JOIN crimes_with_zips AS z 
      ON c.DR_NO = z.DR_NO 
      AND c.LAT = z.LAT 
      AND c.LON = z.LON 
      AND c.X = z.X 
      AND c.year = z.year
    WHERE c.\"Crm.Cd.Desc\" = 'CRIMINAL HOMICIDE'
  ) AS x INNER JOIN (SELECT m.S1903_C03_001E, CAST(m.year AS INTEGER) AS year, CAST(m.ZCTA AS INTEGER) AS ZCTA FROM cali_median_income m) AS i 
    ON x.ZCTA5CE20 = i.ZCTA AND x.year = i.year
")

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}
library(broom)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}
lm_income_crimes <- lm(HomicideCount ~ cali_median_income, data = homicide_income_by_year)

@@

```{r, warning=FALSE, message=FALSE, echo=FALSE, fig.align = "center"}
library(kableExtra)
query <- dbSendQuery(con, "SELECT * FROM crimes_by_census_block")
crimes_by_cb <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}
homicides_latlon_rounded <- homicides %>%
  mutate(LAT_rounded = round(LAT, 2),
         LON_rounded = round(LON, 2))

@@

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align = "center"}

@@

```{r}
knitr::opts_chunk$set(eval = FALSE)
```

```{r, output=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses
library(dplyr)
library(RSQLite)

@@

```{r echo=FALSE, fig.cap="Crime densities in Los Angeles by LAPD Division, message=FALSE, warning=FALSE, 2010.2019 and 2020.Present", fig.align="center"}
query <- dbSendQuery(con, "Select * From crime_la Where year < 2020")
crime_la_2010_2019 <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}
query <- dbSendQuery(con, "Select Count(*) as Crimes, YearWeek From crime_la Group By YearWeek Order By YearWeek")
weekly_counts <- dbFetch(query,n=-1)
dbClearResult(query)

>>>>

We will also investigate the relationship between homicide and enforcement factors. In particular, we will look at the LAPD budget in comparison to the homicide rate to determine if increased police funding is effective in reducing the number of homicides.
```{r, echo = FALSE}
knitr::opts_chunk$set(eval = TRUE, echo = FALSE)
```

```{r DATA, output=FALSE,  message=FALSE, warning=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses
library(dplyr)
library(RSQLite)

@@

```{r  message=FALSE, warning=FALSE, fig.align="center", fig.cap="Crime densities in Los Angeles by LAPD Division, 2010-2019 and 2020-Present"}
query <- dbSendQuery(con, "Select * From crime_la Where year < 2020")
crime_la_2010_2019 <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r,  warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}

query <- dbSendQuery(
  con,
  "
  SELECT
    YearWeek,
    MIN(Date) AS start_of_week,
    MAX(Date) AS end_of_week,
    COUNT(*)  AS Crimes
  FROM crime_la
  GROUP BY YearWeek
  ORDER BY YearWeek
  "
)
weekly_counts <- dbFetch(query, n = -1)
dbClearResult(query)

weekly_counts$start_of_week <- as.Date(weekly_counts$start_of_week)
weekly_counts$end_of_week   <- as.Date(weekly_counts$end_of_week)

weekly_counts$days_in_week <-
  as.numeric(difftime(weekly_counts$end_of_week,
                      weekly_counts$start_of_week,
                      units = "days")) + 1

weekly_counts <- subset(weekly_counts, days_in_week == 7)

weekly_counts$CrimesScaled <-
  weekly_counts$Crimes * (7 / weekly_counts$days_in_week)

weekly_counts$Index <- seq_len(nrow(weekly_counts))

march_2020_index <- which(weekly_counts$YearWeek == "2020-09")

plot(
  weekly_counts$Index,
  weekly_counts$CrimesScaled,
  type  = "l",
  col   = "blue",
  lwd   = 2,
  xlab  = "Weekly Index",
  ylab  = "Number of Crimes (Scaled to 7 Days)"
)

abline(
  h   = pretty(weekly_counts$CrimesScaled),
  v   = pretty(weekly_counts$Index),
  col = "gray",
  lty = "dotted"
)

abline(v = weekly_counts$Index[march_2020_index], col = "red", lwd = 2)

text(
  x     = weekly_counts$Index[march_2020_index] - 80,
  y     = 10,
  labels = "March 2020",
  pos   = 3,
  col   = "red"
)

@@

```{r,  fig.cap = "Los Angeles Crimes by Day of the Week", warning=FALSE, message=FALSE, fig.align = "center"}
# Convert Date.Rptd to Date format (ignore time part)
#crime_la$Date.Rptd
crime_la <- rbind(crime_la_2010_2019, crime_la_2020_present)
# Extract the day of the week
crime_la$Date <- as.Date(substr(crime_la$Date.Rptd, 1, 10), format = "%m/%d/%Y")

@@

```{r,  warning=FALSE, message=FALSE, fig.cap="Percent of crime victims by gender, 2010 to present", fig.align = "center"}
query <- dbSendQuery(con, "SELECT COUNT(*) As count, Modified_Vict_Sex As `Vict.Sex` FROM (
    SELECT *,  
        CASE 
            WHEN `Vict.Sex` NOT IN ('M', 'F') THEN 'N'
            ELSE `Vict.Sex`
        END AS `Modified_Vict_Sex`
    FROM crime_la c
    WHERE `Vict.Sex` != '' AND `Vict.Sex` IS NOT NULL) s
    GROUP BY Modified_Vict_Sex
")
gender_counts <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r,  warning=FALSE, message=FALSE, fig.align = "center"}
zcta <- read_sf(dsn = "0_Data/Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
query <- dbSendQuery(con, "SELECT * FROM zctas_in_la")
LA_zctas <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r,  warning=FALSE, message=FALSE, fig.cap="Median household income by year and by ZIP Code Tabulation Area (ZCTA) in Los Angeles", fig.align = "center"}
# quantiles for the entire dataset (from 2018 to 2023) to divide the median income into groups
qntls <- round(quantile(zcta_no_NAs$S1903_C03_001E, probs = seq(0, 1, length.out = 7)), 0)

@@

```{r,  warning=FALSE, message=FALSE, fig.cap="Annual Median Income and Homicide Count by ZCTA in Los Angeles with top 1% income and top 1% homicide rates labelled", fig.align = "center"}
zcta <- read_sf(dsn = "0_Data/Raw_Data/cb_2020_us_zcta520_500k", layer = "cb_2020_us_zcta520_500k")
query <- dbSendQuery(con, "SELECT i.*, c.*
FROM cali_median_income i
INNER JOIN crimes_with_zips c
ON i.year = c.year AND CAST(i.ZCTA AS int) = c.ZCTA5CE20")
crime_income <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r,  warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Monthly Frequency of Evictions and Crimes in the City of Los Angeles, February 2023 to February 2025"}

@@

```{r,  warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Annual Number of Crimes by Los Angeles Police Department Bureau"}
crime_la$year <- lubridate::year(crime_la$DateParsed)
crime_by_area <- crime_la %>% group_by(AREA.NAME, year) %>% summarize(numberOfCrimes = n()) %>% ungroup()

@@

```{r,  warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Strength of Correlations Between the Number of Crimes and Different Law Enforcement and Socioeconomic Factors"}
library(lubridate)
# creating dataset for correlations

@@

correlations_df <- crime_la_monthly %>% inner_join(unemployment_lapd_budget_evictions, join_by("year" == "Year", "month" == "month")) %>% dplyr::select(month, year, num_evictions, unemployment_rate, num_crimes, police_budget, Crm.Cd.Desc)

@@

write.csv(correlations, "corr.csv")

@@

```{r,  warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Homicide Victims by Age and Gender in Los Angeles 2010-Present"}
homicide_by_gender <- crime_la %>% filter(Crm.Cd.Desc == "CRIMINAL HOMICIDE") %>% dplyr::select(Vict.Sex, DateParsed, Vict.Age) %>% mutate(gender = factor(Vict.Sex, levels  = c("F", "M", "X"), labels = c("Female", "Male", "Other"))) %>% drop_na()

@@

```{r,  warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Frequency of Crimes in Los Angeles Throughout the Day"}
crime_la$time <- sprintf("%04d",crime_la$TIME.OCC)
crime_la$Hour <- as.numeric(substr(as.character(crime_la$time), start = 1, stop = 2))

@@

```{r,  warning=FALSE, message=FALSE, fig.align = "center", fig.cap="Time series analyzing homicide rate trends over time by median income quartiles"}
query <- dbSendQuery(con, "
  SELECT i.S1903_C03_001E, x.year, i.ZCTA, x.LAT, x.LON
  FROM (
    SELECT c.LAT, c.LON, CAST(z.ZCTA5CE20 AS INTEGER) AS ZCTA5CE20, CAST(c.year AS INTEGER) AS year
    FROM crime_la AS c
    INNER JOIN crimes_with_zips AS z 
      ON c.DR_NO = z.DR_NO 
      AND c.LAT = z.LAT 
      AND c.LON = z.LON 
      AND c.X = z.X 
      AND c.year = z.year
    WHERE c.\"Crm.Cd.Desc\" = 'CRIMINAL HOMICIDE'
  ) AS x INNER JOIN (SELECT m.S1903_C03_001E, CAST(m.year AS INTEGER) AS year, CAST(m.ZCTA AS INTEGER) AS ZCTA FROM cali_median_income m) AS i 
    ON x.ZCTA5CE20 = i.ZCTA AND x.year = i.year
")

@@

```{r,  warning=FALSE, message=FALSE, fig.align = "center"}
library(broom)

@@

```{r,  warning=FALSE, message=FALSE, fig.align = "center"}
lm_income_crimes <- lm(HomicideCount ~ cali_median_income, data = homicide_income_by_year)

@@

```{r, warning=FALSE, message=FALSE,  fig.align = "center"}
library(kableExtra)
query <- dbSendQuery(con, "SELECT * FROM crimes_by_census_block")
crimes_by_cb <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r,  warning=FALSE, message=FALSE, fig.align = "center"}
homicides_latlon_rounded <- homicides %>%
  mutate(LAT_rounded = round(LAT, 2),
         LON_rounded = round(LON, 2))

@@

write.csv(homicides_lat_long_yr, "homicides.csv")

@@

```{r,  warning=FALSE, message=FALSE, fig.align = "center"}

@@

```{r, echo = FALSE}
knitr::opts_chunk$set(eval = FALSE, echo = TRUE)
```

```{r, output=FALSE,  message=FALSE, warning=FALSE}
library(sf) # for map
library(tidyverse) # using only for joining datasets to build map visual
library(RColorBrewer) # for map color scheme
library(tidygeocoder) # to get zip codes of addresses
library(dplyr)
library(RSQLite)

@@

```{r  fig.cap="Crime densities in Los Angeles by LAPD Division, message=FALSE, warning=FALSE, 2010.2019 and 2020.Present", fig.align="center"}
query <- dbSendQuery(con, "Select * From crime_la Where year < 2020")
crime_la_2010_2019 <- dbFetch(query,n=-1)
dbClearResult(query)

@@

```{r,  warning=FALSE, message=FALSE, fig.cap = "Weekly Crime Counts in Los Angeles from 2010 to present", fig.align = "center"}
query <- dbSendQuery(con, "Select Count(*) as Crimes, YearWeek From crime_la Group By YearWeek Order By YearWeek")
weekly_counts <- dbFetch(query,n=-1)
dbClearResult(query)
